Phylogeographic inference with microtellite and SNPs

### 2019-07-10 code/rmarkdown/Figures.rmd
map ancestries

>>>>

##Figure 5

??*to decide between top or bottom panel*??
*Top*. Median (horizontal lines) and 25% and 75% quantiles (grey boxes) of the bootstrap support for all nodes in the tree (top panel) from the original datasets and from the subsampled loci. *Bottom*. Bootstrap support for trees from subsampled SNPs from *P. cultripes* (blue dots) and fitted values (solid line) from a linear model on the natural logarithm of the number of loci (R^2 = 0.34; p-value = < 0.001).

##Figure 6

??why are there transparent pies on K8. check code and export as bitmap??
Spatial distribution of localities sampled with pies representing averaged proportion of inferred ancestries from K = 2 to K = 8, for SNP and microsatellite data, in STRUCTURE. The shaded area represent the distributions of *H. molleri* and *P. cultripes*.

### 2019-07-10 code/rmarkdown/article.Rmd
map ancestries

### 2.2.2 Data filtering
We additionally applied several filtering steps to the genotype matrices in R 3.5.1 (R Core Team 2018), using functions from the *dartR* 1.1.11 package (Gruber and Georges 2019) and custom functions (https://github.com/csmiguel/usat_snp; Figure S1). The genotypes were imported into R and filters were applied following order. A per-sample call rate threshold (proportion loci with calls) of 0.35, RepAvg (a repeatability measure of genotypes calls between replicas) threshold of 0.95. We used thresholds for allele balance of 0.15 and 0.85, and removed loci for which their coverage was 3.5 times higher than the median coverage (O'Leary et al. 2018), after which, we filtered loci with a call rate (proportion of samples with a call) below 0.8, retained only one SNP per contig, a set minor allele frequency threshold of 0.02 and removed loci whose alleles deviated from Hardy-Weinberg proportions(O'Leary et al. 2018).
To make the datasets more comparable (Figure 1), we filtered out samples from spatially distant localities between marker datasets.

@@

## 2.4 Phylogenetic structure

We evaluated the phylogenetic relationships between the populations using the neighbor-joining (NJ) algorithm (Saitou and Nei 1987). We discarded using model-based phylogenetic inference for the implicit difficulties of fitting evolutionary and demographic models for intra-species trees (), and for the possibilities that offer comparing the 2 marker types (SNPs and microtellites) under a similar tree-reconstruction method. The application of NJ relies on a distance matrix computed from the genotypes. For the microtellite datasets we computed these matrices using Da distance (Nei et al. 1983) with *genet.dist* function from *hierfstat* 0.04-22 package (Goudet and Jombart 2015) in R. Da distances have been shown to yield more reliable topologies to other distances (Takezaki and Nei 1996, 2008). For the SNP genotypes, we computed euclidean distances with *stats::dist* function in R. Then NJ trees were reconstructed with *njs* function from *ape* 5.0 (Paradis and Schliep 2018) for each dataset. We computed the bootstrap support with a modified approach from *ape::boot.phylo* to accommodate non-standard input data. First, we resampled with replacement loci in the genotype matrices, from which we recomputed NJ trees. These trees were rooted using midpoint rooting with function *midpoint.root* from *phytools* 0.6-60 (Revell 2012) in R. And finally, bootstrap values were computed by counting the partitions of the original trees present in 1000 bootstrapped trees. We inspected the phylogeographic congruence between STRUCTURE and the phylogenetic by plotting together a the consensus solution for the major cluster from the output of CLUMPAK for K = 4, since K = 4, seemed to be an optimal solution for many of the datasets (Figure 2).
*Marker type and bootstrap support*. For each tree, branch lengths were normalized by dividing edge lengths by the length between the most distant tips in the tree. Then, we used several functions in *treeman* 1.1.3 package (Bennett 2018) in R, to extract 2 metrics from each node in the trees: 1) the distance from each node to the root (Dnr), and 2) the distance from each node to its parent node (Dnp). ??For the nodes in each tree we assessed the relation of Dnr and Dnp against their corresponding bootstrap support values fitting generalized linear models (GMLs) using a logit link function, with *stats::glm* function in R. *Actually the fit is really bad. See script 99_04_model_fitting.R and ask Jeff*??

*Number of markers and bootstrap support*.

*MAF and bootstrap support*.

@@

## 3.1 Model-based clustering
*Convergence*. Convergence of the chains measured by the GR diagnostic varied between species and markers. Often, values below 1.05 indicate good convergence (Vats and Knudson 2018 Jeff should check this ref). For SNPs, the chains from each K in STRUCTURE converged well up to K = 2 in *H. molleri* and up to K = 6 in the case of *P. cultripes*. Reducing the number of loci did not affect the pattern of convergence across K in a consistent manner (Figures S2.1, S2.2, S2.5 - S2.11; Figure S3.1).
For microsatellite data, the chains converged consistently up to K = 6 in both species, although the GR diagnostic slightly peaked for K = 5 in *P. cultripes* (Figures S2.3 - S2.4; Figure S3.1).
*Optimal K*. The most optimal number of populations was evaluated with 3 metrics. Pr[X|K] increased steadily from K = 2 to K = 8 in all datasets (Figure 2 A-C). Evanno's ΔK peaked at K = 2 in all SNP datasets, followed by K = 3 (Figure 2 D-F). However, in the microsatellite dataset from *H. molleri*, ΔK peaked at K = 4 and K = 6 (Figure 2D). The predominant K = 2 in the SNP datasets were not in agreement with results from the Parsimony index (Figure 2 G-I), except for the microtellite data in *H. molleri*, in which PI and ΔK peaked at K = 4 (Figure 2G). For SNPs in *H. molleri*, the highest values of the PI were at K = 4 and K = 5, contrary to the K = 2 from ΔK (Figure 2G). In *P. cultripes*, the PI peaked at K = 4 for both markers, and then sunk in the SNP data, whereas for  microtellites the PI reached its maxima at K = 6 - K = 8 (Figure 2H). In the subsets of SNPs from *P. cultripes* the PI increased gradually up to K = 4. After that, the distinct marker sets showed inconsistent trends (Figure 2I).
*clumpak table*

## 3.2 Phylogenetic structure
*Coherence between model-based and NJ clustering*. The phylogeographic structure of the model-based ancestries for the SNP dataset were coherent with the topology of their phylogenetic trees (Figure 4 A-B). That is, the samples from the same populations clustered together on the same or close clades and the ancestries from the samples were coherent with the phylogenetic relationships at different depths in the tree. However, the topology of the trees for microtellites were little supported. Furthermore, it neither provided a solid grouping of the samples from the same populations and it was not able to explain the relationships at different depths in the trees (Figure 4 C-D).

@@

*Number of markers and bootstrap support*.

*MAF and bootstrap support*.

@@

* Overview

* Key result 1

* Key result 2
The little bootstrap support we describe in phylogenetic trees from microtellite data is not surprising, given that often several tens of microsatellites or more are needed to recover the correct topology of trees (Nei et al. 1983; Takezaki and Nei 1996)

* Key result 3

* Concluding paragraph

@@

TODO

>>>>

### 2.2.2 Data filtering
We additionally applied several filtering steps to the genotype matrices in R 3.5.1 (R Core Team 2018), using functions from the *dartR* 1.1.11 package (Gruber and Georges 2019) and custom functions (https://github.com/csmiguel/usat_snp; Figure S1). The genotypes were imported into R and filters were applied following order. A per-sample call rate threshold (proportion loci with calls) of 0.35, RepAvg (a repeatability measure of genotypes calls between replicas) threshold of 0.95. We used thresholds for allele balance of 0.15 and 0.85, and removed loci for which their coverage was 3.5 times higher than the median coverage (O'Leary et al. 2018), after which, we filtered loci with a call rate (proportion of samples with a call) below 0.8, retained only one SNP per contig, a set minor allele frequency threshold (MAF) of 0.02 and removed loci whose alleles deviated from Hardy-Weinberg proportions(O'Leary et al. 2018).
To make the datasets more comparable (Figure 1), we filtered out samples from spatially distant localities between marker datasets.

@@

## 2.4 Phylogenetic structure
We evaluated the phylogenetic relationships between the populations using the neighbor-joining (NJ) algorithm (Saitou and Nei 1987). We discarded using model-based phylogenetic inference for the implicit difficulties of fitting evolutionary and demographic models for intra-species trees (), and for the possibilities that offer comparing the 2 marker types (SNPs and microtellites) under a similar tree-reconstruction method. The application of NJ relies on a distance matrix computed from the genotypes. For the microtellite datasets we computed these matrices using Da distance (Nei et al. 1983) with *genet.dist* function from *hierfstat* 0.04-22 package (Goudet and Jombart 2015) in R. Da distances have been shown to yield more reliable topologies to other distances (Takezaki and Nei 1996, 2008). For the SNP genotypes, we computed euclidean distances with *stats::dist* function in R. Then NJ trees were reconstructed with *njs* function from *ape* 5.0 (Paradis and Schliep 2018) for each dataset. We computed the bootstrap support with a modified approach from *ape::boot.phylo* to accommodate non-standard input data. First, we resampled with replacement loci in the genotype matrices, from which we recomputed NJ trees. These trees were rooted using midpoint rooting with function *midpoint.root* from *phytools* 0.6-60 (Revell 2012) in R. And finally, bootstrap values were computed by counting the partitions of the original trees present in 1000 bootstrapped trees. We inspected the phylogeographic congruence between STRUCTURE and the phylogenetic by plotting together a the consensus solution for the major cluster from the output of CLUMPAK for K = 4, since K = 4, seemed to be an optimal solution for many of the datasets (Figure 2).
*Marker type and bootstrap support*. For each tree, branch lengths were normalized by dividing edge lengths by the length between the most distant tips in the tree. Then, we used several functions in *treeman* 1.1.3 package (Bennett 2018) in R, to extract 2 metrics from each node in the trees: 1) the distance from each node to the root (Dnr), and 2) the distance from each node to its parent node (Dnp). ??For the nodes in each tree we assessed the relation of Dnr and Dnp against their corresponding bootstrap support values fitting generalized linear models (GMLs) using a logit link function, with *stats::glm* function in R. *Actually the fit is really bad. See script 99_04_model_fitting.R and ask Jeff*??

*Number of markers and bootstrap support*. The number of SNPs can affect the phylogeographic patterns inferred from RAD-seq data (Hodel et al. 2017). We evaluated the effect of the number of SNPs on the support of the tree reconstructions. We sampled random subsets of 20000, 10000, 5000, 3000, 1000, 500 and 200 loci from the SNP dataset of *P. cultripes*. This process was repeated 5 times to overcome possible biases on marker choice. On each of these 5 subsets we computed 200 bootstrapped trees as above. We calculated the bootstrap supports by counting the number of partitions of the NJ tree from the full dataset (see above, `r adegenet::nLoc(gen$dart_pelo)` loci) the were present in the 1000 bootstrapped trees. The number of loci was linearized applying a natural logarithm and its relation with the bootstrap support was evaluated with a linear model in R.

*MAF and bootstrap support*. Loci can contain different amount of information. A way to quantify this information is by the MAF (ref). MAF thresholds used in filtering steps can greatly affect population genetic inferences, since allele frequencies are related to the mutation rate of their locus, thus  yielding information about different evolutionary scales (Linck and Battey 2019). Indeed, singletons blur the results of model-based genetic structure analysis (Linck and Battey 2019) and stringent MAF threshold (ie. 0.10) seem to yield best results (Haasl and Payseur 2011). However, their possible effect on the evolutionary relationships between populations has not been studied. We generated SNP datasets by applying MAF thresholds of 0.02, 0.04, 0.06, 0.08 and 0.10, to the SNPs from *P. cultripes*. From each dataset we sampled 500 random loci 5 times and computed bootstrap support from the resulting trees as above. We chose 500 loci because for this particular dataset it is a number of loci sensitive to changes on the overall bootstrap support of the tree (see Results). The relation between the bootstrap supports and MAF filters were assessed with a linear model in R.

@@

## 3.1 Model-based clustering
*Convergence*. Convergence of the chains measured by the GR diagnostic varied between species and markers. Often, values below 1.05 indicate good convergence (Vats and Knudson 2018 Jeff should check this ref). For SNPs, the chains from each K in STRUCTURE converged well up to K = 2 in *H. molleri* and up to K = 6 in the case of *P. cultripes*. Reducing the number of loci did not affect the pattern of convergence across K in a consistent manner (Figures S2.1, S2.2, S2.5 - S2.11; Figure S3.1).
For microsatellite data, the chains converged consistently up to K = 6 in both species, although the GR diagnostic slightly peaked for K = 5 in *P. cultripes* (Figures S2.3 - S2.4; Figure S3.1).
*Optimal K*. The most optimal number of populations was evaluated with 3 metrics. Pr[X|K] increased steadily from K = 2 to K = 8 in all datasets (Figure 2 A-C). Evanno's ΔK peaked at K = 2 in all SNP datasets, followed by K = 3 (Figure 2 D-F). However, in the microsatellite dataset from *H. molleri*, ΔK peaked at K = 4 and K = 6 (Figure 2D). The predominant K = 2 in the SNP datasets were not in agreement with results from the Parsimony index (Figure 2 G-I), except for the microtellite data in *H. molleri*, in which PI and ΔK peaked at K = 4 (Figure 2G). For SNPs in *H. molleri*, the highest values of the PI were at K = 4 and K = 5, contrary to the K = 2 from ΔK (Figure 2G). In *P. cultripes*, the PI peaked at K = 4 for both markers, and then sunk in the SNP data, whereas for  microtellites the PI reached its maxima at K = 6 - K = 8 (Figure 2H). In the subsets of SNPs from *P. cultripes* the PI increased gradually up to K = 4. After that, the distinct marker sets showed inconsistent trends (Figure 2I).
??*clumpak table* talk about major/minor clusters in clumpak or similarities betweend ancestry matrices between mean population ancestries for usat/snp data or between n solutions for K??
For K = 2, the genetic structure was similar for microtellite and SNP data from both species (Figure 6), differentiating a norther and southern lineage for *H. molleri*, and a central-western and north-eastern linage for *P. cultripes*. However, from K = 3 to K = 8, the spatial patterns of genetic structure for both species showed moderate disagreements between SNPs versus microsatellites (Figure 6), which increased with higher values of K. These differences were more marked in *H. molleri*. For instance, in *H. molleri* the southeastern population of Ojos de Villaverde stood out as very distinct from K = 4 from SNPs whereas microtellites detected this singularity at K = 5.??should I focus on the biological patterns to allow a discussion on the phylogegrphy for these species or just on the marker comparison??

## 3.2 Phylogenetic structure
*Coherence between model-based and NJ clustering*. The phylogeographic structure of the model-based ancestries for the SNP dataset were coherent with the topology of their phylogenetic trees (Figure 4 A-B). That is, the samples from the same populations clustered together on the same or close clades and the ancestries from the samples were coherent with the phylogenetic relationships at different depths in the tree. However, the topology of the trees for microtellites were little supported. Furthermore, it neither provided a solid grouping of the samples from the same populations and it was not able to explain the relationships at different depths in the trees (Figure 4 C-D).??look at where the pops are located and talk about the spatial coherence of the phylogegraphic patterns. what can this info provides us regarding the phylogeography of these species respect to microtellites??

@@

*Number of markers and bootstrap support*. The bootstrap support was highest for the dataset with most loci (n = 33144), for which most nodes had a support equal to 1 (Figure 5). The bootstrap support decreased exponentially as the number of markers decreased down to 200 (Figure 5). The relation between bootstrap support and natural logarithm of number of loci was well explained by a linear model (R^2 = 0.34; p-value = < 0.001). The overall BS for the tree from the SNP dataset with lowest number of loci (n = 200 loci) was comparable to the BS for tree from the microtellite data of *H. molleri* (Figure 5).

*MAF and bootstrap support*. There was no effect of the MAF thresholds on the bootstrap support from the trees (t-test = 0.68, p-value = 0.5; Figure S4)

@@

@@
